---
title: "Models for set 3, Simulated F2 results (GAMM modelling strategies)"
author: "Márton Sóskuthy"
date: "12/05/2020"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document is a supplement to "Evaluating generalised additive mixed modelling strategies for dynamic speech analysis," relating specifically to the contents of Table 3 in Section 3.3.2. It presents code that illustrates (i) how the simulated data were generated and (ii) the models whose performance is summarised Table 3.

## Preliminaries

The code below loads the relevant libraries.

```{r, message=F}
library(ggplot2)
library(mgcv)
library(itsadug)
library(MASS)
library(lme4)
```

## Data generation

The code in this section can be used to create data for either type I or type II simulations. Set the value of *type* to 1 for type I simulations and to 2 for type II simulations.

```{r}
type = 2
```

The data for this set of simulations consist of simulated f2 trajectories modelled after the diphthong /aI/. In this simulation, there are 50 different speakers, each represented by 40 trajectories. 20 of these trajectories are assigned to group A and 20 to group B - in other speakers, these simulations exemplify a within-item design. For type I simulations, there is no underlying difference between the two groups. For type II simulations, the underlying trajectories are slightly different (cf. Section 2.1 in the paper and also the Appendix).

The following code sets the parameters that determine the main characteristics of the data set.

```{r}
# setting time dimension
xs = seq(0,1,0.1)

# population parameters: individual speakers come from this dist
f2_start_mean = 1300
if (type==1) {
  f2_end_1_mean = 1630
  f2_end_2_mean = 1660
} else {
  f2_end_1_mean = 1650
  f2_end_2_mean = 1650
}
f2_start_sd.speaker = 90
f2_end_sd.speaker = 90
# correlation between pairs of parameter values sampled for two groups within same speaker;
# rho = 7/8 means that the sd of the difference between the two groups within
#   the same speaker is 0.5 * the sd of the target values
# in mixed model lingo: the sd of the random slope is half of the sd of the intercept
f2_start_rho.speaker = 7/8
f2_end_rho.speaker = 7/8

# expected value & sd for transition point
x0_mean = 0.6
x0_sd.speaker = 0.020
x0_rho.speaker = 7/8 # same as above
# expected value & sd for steepness (higher -> more steep)
k_mean = 15
k_sd.speaker = 4
k_rho.speaker = 7/8 # same as above

# how much variation within speaker-group pairs? (unchanged from before)
f2_start_sd.traj = 150
f2_end_sd.traj = 150
x0_sd.traj = 0.015
k_sd.traj = 3

# amount of random noise

noise_sd <- 40

n_speakers <- 50
n_trajectories_per_speaker <- 40
```

The code below assembles the data set.

```{r}
# assembling trajectories

make_cov_matrix <- function (a.sd, b.sd, rho) {
  matrix(c(a.sd**2, rho*a.sd*b.sd, rho*a.sd*b.sd, b.sd**2), nrow=2)
}

ys_m <- matrix(0, nrow=length(xs), ncol=n_speakers*n_trajectories_per_speaker)
for (i in 1:n_speakers) {
  f2_start.speaker <- mvrnorm(1, rep(f2_start_mean, 2), 
                           make_cov_matrix(f2_start_sd.speaker, f2_start_sd.speaker, f2_start_rho.speaker))
  f2_end.speaker <- mvrnorm(1, c(f2_end_1_mean, f2_end_2_mean), 
                         make_cov_matrix(f2_end_sd.speaker, f2_end_sd.speaker, f2_end_rho.speaker))
  x0.speaker <- mvrnorm(1, rep(x0_mean, 2), 
                     make_cov_matrix(x0_sd.speaker, x0_sd.speaker, x0_rho.speaker))
  k.speaker <- mvrnorm(1, rep(k_mean, 2), 
                     make_cov_matrix(k_sd.speaker, k_sd.speaker, k_rho.speaker))
  for (j in 1:(n_trajectories_per_speaker/2)) {
    # group A
    f2_start <- rnorm(1, f2_start.speaker[1], f2_start_sd.traj)
    f2_end <- rnorm(1, f2_end.speaker[1], f2_end_sd.traj)
    x0 <- rnorm(1, x0.speaker[1], x0_sd.traj)
    k <- rnorm(1, k.speaker[1], k_sd.traj)
    ys_m[,(i-1)*n_trajectories_per_speaker + j*2 - 1] <- ((f2_end - f2_start) / (1 + exp(-k*(xs-x0)))) + f2_start + rnorm(length(xs), 0, noise_sd)
    # group B
    f2_start <- rnorm(1, f2_start.speaker[2], f2_start_sd.traj)
    f2_end <- rnorm(1, f2_end.speaker[2], f2_end_sd.traj)
    x0 <- rnorm(1, x0.speaker[2], x0_sd.traj)
    k <- rnorm(1, k.speaker[2], k_sd.traj)
    ys_m[,(i-1)*n_trajectories_per_speaker + j*2] <- ((f2_end - f2_start) / (1 + exp(-k*(xs-x0)))) + f2_start + rnorm(length(xs), 0, noise_sd)
  }
}

# assembling data set (randomly assigned to categories)
dat <- data.frame(traj=paste("traj_", rep(1:(n_speakers*n_trajectories_per_speaker), each=length(xs)), sep=""),
                  speaker=paste("speaker_", rep(1:n_speakers, each=length(xs)*n_trajectories_per_speaker), sep=""),
                  group=rep(c("A","B"), each=length(xs), times=n_speakers*n_trajectories_per_speaker / 2),
                  measurement.no=xs, 
                  f2=c(ys_m),
                  stringsAsFactors = F
                 )

# setting up different types of grouping factors
dat$group.factor <- as.factor(dat$group)
dat$group.ordered <- as.ordered(dat$group)
contrasts(dat$group.ordered) <- "contr.treatment"
dat$group.bin <- as.numeric(dat$group.factor) - 1

# ids ought to be factors  
dat$traj <- as.factor(dat$traj)
dat$speaker <- as.factor(dat$speaker)
dat$speakerGroup <- interaction(dat$speaker, dat$group)

# add dat$start for AR.start (for autoregressive error models)

dat$start <- dat$measurement.no == 0
```

Here is what the data set looks like. (Note that the difference is very subtle for type II simulations, and hard to make out in this graph. This is because otherwise power would be far too high for the simulations.)

```{r}
ggplot(dat, aes(x=measurement.no, y=f2, group=traj, col=group)) +
  geom_line(alpha=0.6) +
  facet_wrap(~speaker) +
  theme_bw()
```

## Models

All the models (and sets of models) from Table 3 are shown below in the same order as in the table. The numbers in the section headers correspond to the row numbers.

### SIMPLE RANDOM SMOOTHS: 1. Random smooth by speaker, 3 bs

```{r}
rand_smooth_3 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 3), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(rand_smooth_3)
```

### SIMPLE RANDOM SMOOTHS: 2. Random smooth by speaker, 5 bs

```{r}
rand_smooth_5 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(rand_smooth_5)
```

### SIMPLE RANDOM SMOOTHS: 3. Random smooth by speaker, 10 bs

```{r}
rand_smooth_10 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 10), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(rand_smooth_10)
```

### RANDOM SMOOTHS + SLOPES: 4. 3 bs

```{r}
rand_slope_smooth_3 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 3) +
        s(speaker, group.ordered, bs="re"), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(rand_slope_smooth_3)
```

### RANDOM SMOOTHS + SLOPES: 5. 5 bs

```{r}
rand_slope_smooth_5 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5) +
        s(speaker, group.ordered, bs="re"), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(rand_slope_smooth_5)
```

### RANDOM SMOOTHS + SLOPES: 6. 10 bs

```{r}
rand_slope_smooth_10 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 10) +
        s(speaker, group.ordered, bs="re"), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(rand_slope_smooth_10)
```

### ITEM X EFFECT: 7. Item x effect random smooths, 3 bs

```{r}
item_x_effect_rand_smooth_3 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speakerGroup, bs = "fs", m = 1, xt = "cr", k = 3), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(item_x_effect_rand_smooth_3)
```

### ITEM X EFFECT: 8. Item x effect random smooths, 5 bs

```{r}
item_x_effect_rand_smooth_5 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speakerGroup, bs = "fs", m = 1, xt = "cr", k = 5), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(item_x_effect_rand_smooth_5)
```

### ITEM X EFFECT: 9. Item x effect random smooths, 10 bs

```{r}
item_x_effect_rand_smooth_10 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speakerGroup, bs = "fs", m = 1, xt = "cr", k = 10), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(item_x_effect_rand_smooth_10)
```

### ITEM BY EFFECT: 10. Item by effect random smooths, 3 bs

```{r}
item_by_effect_rand_smooth_3 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, by=group.factor, bs = "fs", m = 1, xt = "cr", k = 3), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(item_by_effect_rand_smooth_3)
```

### ITEM BY EFFECT: 11. Item by effect random smooths, 5 bs

```{r}
item_by_effect_rand_smooth_5 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, by=group.factor, bs = "fs", m = 1, xt = "cr", k = 5), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(item_by_effect_rand_smooth_5)
```

### ITEM BY EFFECT: 12. Item by effect random smooths, 10 bs

```{r}
item_by_effect_rand_smooth_10 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, by=group.factor, bs = "fs", m = 1, xt = "cr", k = 10), 
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(item_by_effect_rand_smooth_10)
```

### REFERENCE/DIFFERENCE: 13. random reference/difference smooths, 3 bs

```{r}
reference_difference_rand_smooth_3 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 3) +
        s(measurement.no, speaker, by=group.ordered, bs = "fs", m = 1, xt = "cr", k = 3),
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(reference_difference_rand_smooth_3)
```

### REFERENCE/DIFFERENCE: 14. random reference/difference smooths, 5 bs

```{r}
reference_difference_rand_smooth_5 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5) +
        s(measurement.no, speaker, by=group.ordered, bs = "fs", m = 1, xt = "cr", k = 5),
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(reference_difference_rand_smooth_5)
```

### REFERENCE/DIFFERENCE: 15. random reference/difference smooths, 10 bs

```{r}
reference_difference_rand_smooth_10 <- 
  bam(f2 ~ group.ordered + 
        s(measurement.no, bs = "tp", k = 10) + 
        s(measurement.no, by = group.ordered, bs = "tp", k = 10) + 
        s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 10) +
        s(measurement.no, speaker, by=group.ordered, bs = "fs", m = 1, xt = "cr", k = 10),
      data = dat, 
      method = "fREML", discrete = T, nthreads = 1)
summary(reference_difference_rand_smooth_10)
```

### An additional mini-type II error simulation for item x effect random effects with lme4

This mini-simulation demonstrates that item x effect random effects lead to decreased power for conventional mixed effects models in the same way that they do for GAMMs.

We first set the parameters for this type II error simulation. We simulate 20 "speakers" who each produce tokens from two groups with slightly different mean values for some phonetic parameter. Tokens from the same speaker are not independent of each other.

```{r}
# population means for tokens a / b
mean.a <- 0
mean.b <- 10

# sd of speaker means around population mean
sd.a <- 40
sd.b <- 40
# correlation within speakers
rho.sim <- 7/8

n.speakers <- 20
resid.sd <- 30 # residual standard deviation
n <- 20

# how many simulation runs?
iterations <- 1000
```

We now run the simulation. The two numbers at the end represent power for (i) a model with appropriately specified random intercepts and slopes and (ii) a model with item x effect random effects. The regular method has much higher power than item x effect random effects.

```{r}
# function for making covariance matrices
make_cov_matrix <- function (a.sd, b.sd, rho) {
  matrix(c(a.sd**2, rho*a.sd*b.sd, rho*a.sd*b.sd, b.sd**2), nrow=2)
}

# vectors for storing results
ts.regular <- c()
ts.itemxeffect <- c()

# running the actual simulation
for (i in 1:iterations) {
  # generating by-speaker means
  means <- mvrnorm(n.speakers, c(mean.a, mean.b), 
                   make_cov_matrix(sd.a, sd.b, rho.sim))
  # generating individual tokens
  d.a <- unlist(lapply(means[,1], function (x) rnorm(n, x, resid.sd)))
  d.b <- unlist(lapply(means[,2], function (x) rnorm(n, x, resid.sd)))
  # assembling data frame
  d <- data.frame(y=c(d.a, d.b),
                group=as.factor(rep(c("a","b"), each=length(d.a))),
                speaker=as.factor(paste("speaker", rep(1:n.speakers, each=n)))
                )
  # adding combined predictor
  d$speakerGroup <- interaction(d$speaker, d$group)
  # fitting regular model
  mod.reg <- suppressWarnings(lmer(y ~ group + (1 + group | speaker),
                  data=d))
  # fitting item x effect model
  mod.ixe <- suppressWarnings(lmer(y ~ group + (1 | speakerGroup),
                  data=d))
  ts.regular <- c(ts.regular, summary(mod.reg)$coefficients[2,3])
  ts.itemxeffect <- c(ts.itemxeffect, summary(mod.ixe)$coefficients[2,3])
  # cat("\r                 \r", i)
}
mean(abs(ts.regular) > 1.96)
mean(abs(ts.itemxeffect) > 1.96)
```