---
title: "Models for set 2, Simulated F2 results (GAMM modelling strategies)"
author: "Márton Sóskuthy"
date: "12/05/2020"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document is a supplement to "Evaluating generalised additive mixed modelling strategies for dynamic speech analysis," relating specifically to the contents of the "simulated formants" columns of Table 4 in Section 3.4.2. It presents code that illustrates (i) how the simulated data were generated and (ii) the models whose performance is summarised in the "simulated formants" columns of Table 4.

## Preliminaries

The code below loads the relevant libraries.

```{r, message=F}
library(ggplot2)
library(mgcv)
library(itsadug)
library(stringr)
library(dplyr)
```

## Data generation

The code in this section can be used to create data for either type I or type II simulations. Set the value of *type* to 1 for type I simulations and to 2 for type II simulations.

```{r}
type = 2
```

The data for this set of simulations consist of simulated f2 trajectories modelled after the diphthong /aI/. 30 speakers are simulated with 40 trajectories each. For type I simulations, each speaker is randomly assigned to one of two groups (A or B). For type II simulations, all underlying trajectories for group B speakers are slightly modified (cf. Section 2.1 in the paper and also the Appendix).

Parameters for the simulations.

```{r}
# setting time dimension
xs = seq(0,1,0.1)

# population parameters: individual speakers come from this dist
f2_start_mean = 1300

if (type==1) {
  f2_end_1_mean = 1650
  f2_end_2_mean = 1650
} else {
  f2_end_1_mean = 1600
  f2_end_2_mean = 1690
}

f2_start_sd.speaker = 90
f2_end_1_sd.speaker = 90
f2_end_2_sd.speaker = 90
# expected value & sd for transition point
x0_mean = 0.6
x0_sd.speaker = 0.020
# expected value & sd for steepness (higher -> more steep)
k_mean = 15
k_sd.speaker = 4

# how much variation within speakers?
f2_start_sd.traj = 150
f2_end_1_sd.traj = 150
f2_end_2_sd.traj = 150
x0_sd.traj = 0.015
k_sd.traj = 3

# amount of random noise

noise_sd <- 40

n_speakers <- 30
n_trajectories_per_speaker <- 40
```

Generating simulated data.

```{r}
# assembling trajectories

ys_m <- matrix(0, nrow=length(xs), ncol=n_speakers*n_trajectories_per_speaker)
for (i in 1:n_speakers) {
  f2_start.speaker <- rnorm(1, f2_start_mean, f2_start_sd.speaker)
  f2_end_1.speaker <- rnorm(1, f2_end_1_mean, f2_end_1_sd.speaker)
  f2_end_2.speaker <- rnorm(1, f2_end_2_mean, f2_end_2_sd.speaker)
  x0.speaker <- rnorm(1, x0_mean, x0_sd.speaker)
  k.speaker <- rnorm(1, k_mean, k_sd.speaker)
  for (j in 1:n_trajectories_per_speaker) {
    f2_start <- rnorm(1, f2_start.speaker, f2_start_sd.traj)
    f2_end_1 <- rnorm(1, f2_end_1.speaker, f2_end_1_sd.traj)
    f2_end_2 <- rnorm(1, f2_end_2.speaker, f2_end_2_sd.traj)
    x0 <- rnorm(1, x0.speaker, x0_sd.traj)
    k <- rnorm(1, k.speaker, k_sd.traj)
    if (i <= (n_speakers / 2)) {
      ys_m[,(i-1)*n_trajectories_per_speaker + j] <- ((f2_end_1 - f2_start) / (1 + exp(-k*(xs-x0)))) + f2_start + rnorm(length(xs), 0, noise_sd)
    } else {
      ys_m[,(i-1)*n_trajectories_per_speaker + j] <- ((f2_end_2 - f2_start) / (1 + exp(-k*(xs-x0)))) + f2_start + rnorm(length(xs), 0, noise_sd)
    }
  }
}

# assembling data set (randomly assigned to categories)
dat <- data.frame(traj=paste("traj_", rep(1:(n_speakers*n_trajectories_per_speaker), each=length(xs)), sep=""),
                        speaker=paste("speaker_", rep(1:n_speakers, each=length(xs)*n_trajectories_per_speaker), sep=""),
                        group=rep(c("A","B"), each=length(xs)*(n_speakers*n_trajectories_per_speaker / 2)),
                        measurement.no=xs, 
                        f2=c(ys_m),
                        stringsAsFactors = F
)

# setting up different types of grouping factors
dat$group.factor <- as.factor(dat$group)
dat$group.ordered <- as.ordered(dat$group)
contrasts(dat$group.ordered) <- "contr.treatment"
dat$group.bin <- as.numeric(dat$group.factor) - 1

# ids ought to be factors  
dat$traj <- as.factor(dat$traj)
dat$speaker <- as.factor(dat$speaker)

# add dat$start for AR.start (for autoregressive error models)

dat$start <- dat$measurement.no == 0

# dat$start has already been added at data prep stage (for AR.start, i.e. for autoregressive error models)
```

Here is what the data set looks like.

```{r}
ggplot(dat, aes(x=measurement.no, y=f2, group=traj, col=group)) +
  geom_line(alpha=0.6) +
  facet_wrap(~speaker) +
  theme_bw()
```

## Methods of significance testing

All the models (and sets of models) from Table 4 are shown below in the same order as in the table. The numbers in the section headers correspond to the row numbers.

### MODEL SUMMARY: 1. Looking at model summary

```{r}
modsum <- bam(f2 ~ group.ordered + 
                      s(measurement.no, bs = "tp", k = 10) + 
                      s(measurement.no, by = group.ordered, bs = "tp", k = 10) +
                      s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5), 
                    data = dat, 
                    method = "fREML", discrete = T, nthreads = 1)
summary(modsum)
```

### MODEL SUMMARY: 2. Looking at model summary + Bonferroni correction

For the Bonferroni correction, the alpha-level of the parametric and smooth terms is lowered to 0.025. This does not require fitting a separate model.

### MODEL SUMMARY: 3. Binary smooth

```{r}
binsmooth <- bam(f2 ~ s(measurement.no, bs = "tp", k = 10) + 
                      s(measurement.no, by = group.bin, bs = "tp", k = 10) +
                      s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5), 
                    data = dat, 
                    method = "fREML", discrete=T, nthreads=1)
summary(binsmooth)
```

### LIKELIHOOD RATIO TESTS: 4. Likelihood Ratio Test using models fitted with ML

Please note that these models may take quite a while to fit (5-10 minutes).

```{r}
lrt_ML_full <- bam(f2 ~ group.ordered + 
                      s(measurement.no, bs = "tp", k = 10) + 
                      s(measurement.no, by = group.ordered, bs = "tp", k = 10) +
                      s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5), 
                    data = dat, 
                    method = "ML")
lrt_ML_nested <- bam(f2 ~ # group.ordered + 
                      s(measurement.no, bs = "tp", k = 10) + 
                      # s(measurement.no, by = group.ordered, bs = "tp", k = 10) +
                      s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5), 
                    data = dat, 
                    method = "ML")
compareML(lrt_ML_full, lrt_ML_nested)
```

### LIKELIHOOD RATIO TESTS: 5. Likelihood Ratio Test using models fitted with fREML

As noted in the main text of the paper, the results of this model comparison are meaningless.

```{r}
lrt_fREML_full <- bam(f2 ~ group.ordered + 
                      s(measurement.no, bs = "tp", k = 10) + 
                      s(measurement.no, by = group.ordered, bs = "tp", k = 10) +
                      s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5), 
                    data = dat, 
                    method = "fREML", discrete=T, nthreads=1)
lrt_fREML_nested <- bam(f2 ~ # group.ordered + 
                      s(measurement.no, bs = "tp", k = 10) + 
                      # s(measurement.no, by = group.ordered, bs = "tp", k = 10) +
                      s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5), 
                    data = dat, 
                    method = "fREML", discrete=T, nthreads=1)
compareML(lrt_fREML_full, lrt_fREML_nested)
```

### LIKELIHOOD RATIO TESTS: 6. Likelihood Ratio Test with fREML trick

For this model, fixed effects are estimated as random effects, which makes model comparison based on models fitted with (f)REML valid in principle.

```{r}
lrt_fREML_trick_full <- bam(f2 ~ s(group.ordered, bs="re") + 
                      s(measurement.no, bs = "tp", k = 10) + 
                      s(measurement.no, by = group.ordered, bs = "tp", k = 10) +
                      s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5), 
                    data = dat, 
                    method = "fREML", discrete=T, nthreads=1,
                    select=T)
lrt_fREML_trick_nested <- bam(f2 ~ # s(group.ordered, bs="re") + 
                      s(measurement.no, bs = "tp", k = 10) + 
                      # s(measurement.no, by = group.ordered, bs = "tp", k = 10) +
                      s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5), 
                    data = dat, 
                    method = "fREML", discrete=T, nthreads=1,
                    select=T)
compareML(lrt_fREML_trick_full, lrt_fREML_trick_nested)
```

### Model comparison with AIC (models fitted with fREML)

As noted in the main text of the paper, the results of this model comparison are meaningless.

```{r}
aic_fREML_full <- bam(f2 ~ group.ordered + 
                      s(measurement.no, bs = "tp", k = 10) + 
                      s(measurement.no, by = group.ordered, bs = "tp", k = 10) +
                      s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5), 
                    data = dat, 
                    method = "fREML", discrete=T, nthreads=1)
aic_fREML_nested <- bam(f2 ~ # group.ordered + 
                      s(measurement.no, bs = "tp", k = 10) + 
                      # s(measurement.no, by = group.ordered, bs = "tp", k = 10) +
                      s(measurement.no, speaker, bs = "fs", m = 1, xt = "cr", k = 5), 
                    data = dat, 
                    method = "fREML", discrete=T, nthreads=1)
AIC(aic_fREML_full, aic_fREML_nested)
```

### VISUAL TESTS

This is not a detailed implementation of the percentage-cut-off-based reasoning examined in the paper, simply some example code that can be used to generate (i) prediction plots with confidence intervals for the two groups and (ii) plots of the estimated difference between the groups. Note also that these pred

```{r}
plot_smooth(modsum, view="measurement.no", plot_all="group.ordered", rm.ranef=T)
plot_diff(modsum, view="measurement.no", comp=list(group.ordered=c("A","B")), rm.ranef=T)
```
