---
title: "GAMM Modelling Strategies paper (JPhon): Analysis"
output: html_notebook
---

```{r "setup", include=FALSE}
require("knitr")
# setting workind dir
opts_knit$set(root.dir = "~/documents/research/projects/dynamic-gam/gamm_strategies")
```

Importing relevant libraries & main data set.

```{r}
library(mgcv)
library(itsadug)
library(data.table)
library(stringr)
library(xtable)
library(tidyverse)

dat <- readRDS("gamm_sim_master.rds")
```

## AR models & random effects for capturing autocorrelation

First, let's have a look at how well different types of random structures and AR models do at capturing autocorrelation within (relatively) smooth trajectories. We'll only look at a subset of the data at this point: `sim.type == "simple"`. We'll go straight for a well-formatted table.

All three data types go in a single table, but type II error rates and processing time / memory info are only provided for the simulated data.

```{r}
dat.simple.all <- dat[dat$sim.type=="simple" & !(AR=="AR_est" & ranef!="noranef"),]
# aggregating data
dat.simple.all.aggr <- 
  dat.simple.all[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                    par.type.1=mean(par.p[error.type=="type_1"] < 0.05),
                    smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                    par.power=mean(par.p[error.type=="type_2"] < 0.05),
                    time.med=median(time),
                    memory.med=median(memory)),
                  by=.(data.type,ranef,AR)]
kable(dat.simple.all.aggr)

dat.simple.all.aggr$ranef <- recode(dat.simple.all.aggr$ranef, 
                                 gamcheck_tp_4_10_2="gamcheck_tp",
                                 gamcheck_cr_4_10_2="gamcheck_cr",
                                 gamcheck_tp_4_16_3="gamcheck_tp",
                                 gamcheck_cr_4_16_3="gamcheck_cr")

# creating output table for formant data
# record timing / memory for dense data, set aside
sf.time.mem <- dat.simple.all.aggr[data.type=="simulated",c("ranef","AR","time.med","memory.med")]
# recode gam.check

# go to wide format
sf.out <- dcast.data.table(dat.simple.all.aggr[!(!ranef=="noranef" & AR=="AR_est"),], ranef + AR ~ data.type, value.var=c("par.type.1","smooth.type.1","par.power","smooth.power"))
# merge back with timing / memory stuff
sf.out <- merge(sf.out, sf.time.mem, by=c("ranef","AR"), all.x=T)
# reorder rows
sf.out <- sf.out[c(4:6,15,16,12:14,2,10,11,7:9,1,3),]
# new column that codes ranef and AR together
sf.out[,model.component:=ifelse(AR=="AR_est","AR1",ranef)]
sf.out$model.component <- recode(sf.out$model.component,
                                 noranef="no components",
                                 rintcpt="r intcpt",
                                 rslope="r intcpt + slope",
                                 rsmooth_tp_3="r smooth, tp, 3 bs",
                                 rsmooth_tp_5="r smooth, tp, 5 bs",
                                 rsmooth_tp_10="r smooth, tp, 10 bs",
                                 rsmooth_tp_15="r smooth, tp, 15 bs",
                                 rsmooth_tp_20="r smooth, tp, 20 bs",
                                 rsmooth_cr_3="r smooth, cr, 3 bs",
                                 rsmooth_cr_5="r smooth, cr, 5 bs",
                                 rsmooth_cr_10="r smooth, cr, 10 bs",
                                 rsmooth_cr_15="r smooth, cr, 15 bs",
                                 rsmooth_cr_20="r smooth, cr, 20 bs",
                                 gamcheck_tp="r smooth, tp, gam.check",
                                 gamcheck_cr="r smooth, cr, gam.check")
sf.out <- dplyr::select(sf.out,
                        -ranef, -AR,
                        -par.power_formants, -par.power_pitch, -par.power_simulated,
                        -smooth.power_formants, -smooth.power_pitch)
setcolorder(sf.out, c("model.component",
                      "par.type.1_formants","smooth.type.1_formants",
                      "par.type.1_pitch","smooth.type.1_pitch",
                      "par.type.1_simulated","smooth.type.1_simulated","smooth.power_simulated",
                      "memory.med","time.med"))
colnames(sf.out) <- c("model component",
                      "type I","type I",
                      "type I","type I",
                      "type I","type I","power",
                      "mem.","time")
sf.out <- as.data.frame(sf.out)
for (i in 2:8) {
  sf.out[,i] <- sprintf("%.3f", sf.out[,i])
}
for (i in 9:10) {
  sf.out[,i] <- sprintf("%.1f", sf.out[,i])
}

for (i in 2:8) {
  if (colnames(sf.out)[i]=="type I") {
    for (j in 1:nrow(sf.out)) {
      if (sf.out[j,i]!="NA" && as.numeric(sf.out[j,i]) > 0.1) {
        sf.out[j,i] <- paste0("\\color{lightgray} ",sf.out[j,i])
        if (colnames(sf.out)[i+1]=="power") {
          sf.out[j,i+1] <- paste0("\\color{lightgray} ",sf.out[j,i+1])
        }
      }
    }
  }
}
for (i in 2:10) {
  sf.out[sf.out[,i]=="NA",i] <- "--"
}
sf.str.out <- print(xtable(sf.out), include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
sf.str.out <- gsub(
  "\\\\begin\\{tabular\\}\\{l*\\}", 
  "\\\\begin{tabular}{lrr@{\\\\hspace{1.8em}}rr@{\\\\hspace{1.8em}}rrr@{\\\\hspace{1.8em}}rr}", sf.str.out)
sf.str.out <- gsub("toprule",
"toprule 
  & \\\\multicolumn{2}{c@{\\\\hspace{1.8em}}}{\\\\textsc{real formants}} & \\\\multicolumn{2}{c@{\\\\hspace{1.8em}}}{\\\\textsc{real pitch}} & \\\\multicolumn{3}{c@{\\\\hspace{1.8em}}}{\\\\textsc{simulated formants}}\\\\\\\\ 
\\\\cmidrule(l{.5em}r{1.7em}){2-3}\\\\cmidrule(l{.5em}r{1.7em}){4-5}\\\\cmidrule(l{.5em}r{1.7em}){6-8}
  & \\\\multicolumn{1}{c}{\\\\textsc{param.}} & \\\\multicolumn{1}{c@{\\\\hspace{1.8em}}}{\\\\textsc{smooth}} &
    \\\\multicolumn{1}{c}{\\\\textsc{param.}} & \\\\multicolumn{1}{c@{\\\\hspace{1.8em}}}{\\\\textsc{smooth}} &
    \\\\multicolumn{1}{c}{\\\\textsc{param.}} & \\\\multicolumn{2}{c@{\\\\hspace{1.8em}}}{\\\\textsc{smooth}} & & \\\\\\\\
\\\\cmidrule(lr{.5em}){2-2}\\\\cmidrule(l{.5em}r{1.7em}){3-3}
\\\\cmidrule(lr{.5em}){4-4}\\\\cmidrule(l{.5em}r{1.7em}){5-5}
\\\\cmidrule(lr{.5em}){6-6}\\\\cmidrule(l{.5em}r{1.7em}){7-8}", sf.str.out)
sf.str.out <- gsub("table","table*",sf.str.out)
clip <- pipe("pbcopy", "w")
write(sf.str.out, file=clip)                               
close(clip)
#cat(str.out)
```

Distribution of k values in gam.check?

```{r}
# formants
hist(unlist(dat.simple.all[startsWith(ranef,"gamcheck") & data.type=="formants" & error.type=="type_1","random.smooth.k"]))
# pitch, thin
hist(unlist(dat.simple.all[startsWith(ranef,"gamcheck") & data.type=="pitch" & error.type=="type_1","random.smooth.k"]))
# simulated formants
hist(unlist(dat.simple.all[startsWith(ranef,"gamcheck") & data.type=="simulated" & error.type=="type_1","random.smooth.k"]))
```


## Random effects for capturing across-group patterns

`sim.type == "complex_across"`.

```{r}
dat.ca.all <- dat[sim.type=="complex_across" & ((AR=="noAR" & data.type=="simulated") | (AR=="AR_est" & data.type!="simulated")),]
# aggregating data
dat.ca.all.aggr <- 
  dat.ca.all[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                    par.type.1=mean(par.p[error.type=="type_1"] < 0.05),
                    smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                    par.power=mean(par.p[error.type=="type_2"] < 0.05),
                    time.med=median(time),
                    memory.med=median(memory)),
                  by=.(data.type,ranef,AR)]
kable(dat.ca.all.aggr)

dat.ca.all.aggr$ranef <- gsub("noranef[+]","", dat.ca.all.aggr$ranef)
dat.ca.all.aggr$ranef <- recode(dat.ca.all.aggr$ranef, 
                                 `gamcheck_tp_4_10_3`="gamcheck_tp",
                                 `gamcheck_cr_4_10_3`="gamcheck_cr",
                                 `gamcheck_tp_4_16_3`="gamcheck_tp",
                                 `gamcheck_cr_4_16_3`="gamcheck_cr")

# creating output table for formant data
# record timing / memory for dense data, set aside
sf.time.mem <- dat.ca.all.aggr[data.type=="simulated",c("ranef","time.med","memory.med")]
# recode gam.check

# go to wide format
sf.out <- dcast.data.table(dat.ca.all.aggr, ranef ~ data.type, value.var=c("par.type.1","smooth.type.1","par.power","smooth.power"))
# merge back with timing / memory stuff
sf.out <- merge(sf.out, sf.time.mem, by=c("ranef"), all.x=T)
# reorder rows
sf.out <- sf.out[c(3:5,12,13,10,11,2,8,9,6,7,1),]
# new column that codes ranef
sf.out[,model.component:=ranef]
sf.out$model.component <- recode(sf.out$model.component,
                                 noranef="no components",
                                 rintcpt="r intcpt",
                                 rslope="r intcpt + slope",
                                 rsmooth_tp_3="r smooth, tp, 3 bs",
                                 rsmooth_tp_5="r smooth, tp, 5 bs",
                                 rsmooth_tp_10="r smooth, tp, 10 bs",
                                 rsmooth_tp_15="r smooth, tp, 15 bs",
                                 rsmooth_cr_3="r smooth, cr, 3 bs",
                                 rsmooth_cr_5="r smooth, cr, 5 bs",
                                 rsmooth_cr_10="r smooth, cr, 10 bs",
                                 rsmooth_cr_15="r smooth, cr, 15 bs",
                                 gamcheck_tp="r smooth, tp, gam.check",
                                 gamcheck_cr="r smooth, cr, gam.check")
sf.out <- dplyr::select(sf.out,
                        -ranef,
                        -par.power_formants, -par.power_pitch, -par.power_simulated,
                        -smooth.power_formants, -smooth.power_pitch)
setcolorder(sf.out, c("model.component",
                      "par.type.1_formants","smooth.type.1_formants",
                      "par.type.1_pitch","smooth.type.1_pitch",
                      "par.type.1_simulated","smooth.type.1_simulated","smooth.power_simulated",
                      "memory.med","time.med"))
colnames(sf.out) <- c("model component",
                      "type I","type I",
                      "type I","type I",
                      "type I","type I","power",
                      "mem.","time")
sf.out <- as.data.frame(sf.out)
for (i in 2:8) {
  sf.out[,i] <- sprintf("%.3f", sf.out[,i])
}
for (i in 9:10) {
  sf.out[,i] <- sprintf("%.1f", sf.out[,i])
}

for (i in 2:8) {
  if (colnames(sf.out)[i]=="type I") {
    for (j in 1:nrow(sf.out)) {
      if (sf.out[j,i]!="NA" && as.numeric(sf.out[j,i]) > 0.1) {
        sf.out[j,i] <- paste0("\\color{lightgray} ",sf.out[j,i])
        if (colnames(sf.out)[i+1]=="power") {
          sf.out[j,i+1] <- paste0("\\color{lightgray} ",sf.out[j,i+1])
        }
      }
    }
  }
}
for (i in 2:10) {
  sf.out[sf.out[,i]=="NA",i] <- "--"
}
sf.str.out <- print(xtable(sf.out), include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
sf.str.out <- gsub(
  "\\\\begin\\{tabular\\}\\{l*\\}", 
  "\\\\begin{tabular}{lrr@{\\\\hspace{1.8em}}rr@{\\\\hspace{1.8em}}rrr@{\\\\hspace{1.8em}}rr}", sf.str.out)
sf.str.out <- gsub("toprule",
"toprule 
  & \\\\multicolumn{2}{c@{\\\\hspace{1.8em}}}{\\\\textsc{real formants}} & \\\\multicolumn{2}{c@{\\\\hspace{1.8em}}}{\\\\textsc{real pitch}} & \\\\multicolumn{3}{c@{\\\\hspace{1.8em}}}{\\\\textsc{simulated formants}}\\\\\\\\ 
\\\\cmidrule(l{.5em}r{1.7em}){2-3}\\\\cmidrule(l{.5em}r{1.7em}){4-5}\\\\cmidrule(l{.5em}r{1.7em}){6-8}
  & \\\\multicolumn{1}{c}{\\\\textsc{param.}} & \\\\multicolumn{1}{c@{\\\\hspace{1.8em}}}{\\\\textsc{smooth}} &
    \\\\multicolumn{1}{c}{\\\\textsc{param.}} & \\\\multicolumn{1}{c@{\\\\hspace{1.8em}}}{\\\\textsc{smooth}} &
    \\\\multicolumn{1}{c}{\\\\textsc{param.}} & \\\\multicolumn{2}{c@{\\\\hspace{1.8em}}}{\\\\textsc{smooth}} & & \\\\\\\\
\\\\cmidrule(lr{.5em}){2-2}\\\\cmidrule(l{.5em}r{1.7em}){3-3}
\\\\cmidrule(lr{.5em}){4-4}\\\\cmidrule(l{.5em}r{1.7em}){5-5}
\\\\cmidrule(lr{.5em}){6-6}\\\\cmidrule(l{.5em}r{1.7em}){7-8}", sf.str.out)
sf.str.out <- gsub("table","table*",sf.str.out)
clip <- pipe("pbcopy", "w")
write(sf.str.out, file=clip)                               
close(clip)
#cat(str.out)
```

Distribution of k values in gam.check?

```{r}
# formants
hist(unlist(dat.ca.all[startsWith(ranef,"noranef+gamcheck") & data.type=="formants" & error.type=="type_1","random.smooth.k"]))
# pitch, thin
hist(unlist(dat.ca.all[startsWith(ranef,"noranef+gamcheck") & data.type=="pitch" & error.type=="type_1","random.smooth.k"]))
# simulated formants
hist(unlist(dat.ca.all[startsWith(ranef,"noranef+gamcheck") & data.type=="simulated" & error.type=="type_1","random.smooth.k"]))
```



## Random effects for capturing within-group patterns

`sim.type == "complex_within"`.

```{r}
dat.cw.formants <- dat[dat$sim.type=="complex_within" & dat$data.type=="simulated",]
# aggregating data
dat.cw.formants.aggr <- 
  dat.cw.formants[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                         par.type.1=mean(par.p[error.type=="type_1"] < 0.05),
                         smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                         par.power=mean(par.p[error.type=="type_2"] < 0.05),
                         time.med=median(time),
                         memory.med=median(memory)),
                      by=.(data.density,ranef)]
kable(dat.cw.formants.aggr)

# creating output table for formant data
# record timing / memory for dense data, set aside
#cwf.time.mem <- dat.cw.formants.aggr[data.density=="dense",c("ranef","time.med","memory.med")]
# go to wide format
cwf.out <- dcast.data.table(dat.cw.formants.aggr, ranef ~ ., value.var=c("par.type.1","smooth.type.1","par.power","smooth.power","time.med","memory.med"))
# reorder rows
cwf.out <- cwf.out[c(2,3,1,14,15,13,5,6,4,11,12,10,8,9,7),]
# new column that codes ranef
cwf.out$ranef <- gsub("noranef[+]","",cwf.out$ranef)
cwf.out$model.component <- recode(cwf.out$ranef,
                                 `rsmooth_cr_3`="cr, 3 bs",
                                 `rsmooth_cr_5`="cr, 5 bs",
                                 `rsmooth_cr_10`="cr, 10 bs",
                                 `rsmoothslope_cr_3`="cr, 3 bs",
                                 `rsmoothslope_cr_5`="cr, 5 bs",
                                 `rsmoothslope_cr_10`="cr, 10 bs",
                                 `rsmoothcrossed_cr_3`="cr, 3 bs",
                                 `rsmoothcrossed_cr_5`="cr, 5 bs",
                                 `rsmoothcrossed_cr_10`="cr, 10 bs",
                                 `rsmoothsep_cr_3`="cr, 3 bs",
                                 `rsmoothsep_cr_5`="cr, 5 bs",
                                 `rsmoothsep_cr_10`="cr, 10 bs",
                                 `rsmoothdiff_cr_3`="cr, 3 bs",
                                 `rsmoothdiff_cr_5`="cr, 5 bs",
                                 `rsmoothdiff_cr_10`="cr, 10 bs"
                                 )
cwf.out <- cwf.out[,-c(1)]
setcolorder(cwf.out, c("model.component",
                      "par.type.1","par.power",
                      "smooth.type.1","smooth.power",
                      "memory.med","time.med"))
colnames(cwf.out) <- c("model component",
                      "type I","power",
                      "type I","power",
                      "memory","time")
cwf.out <- as.data.frame(cwf.out)
for (i in 2:5) {
  cwf.out[,i] <- sprintf("%.3f", cwf.out[,i])
}
for (i in 6:7) {
  cwf.out[,i] <- sprintf("%.1f", cwf.out[,i])
}

for (i in 2:5) {
  if (colnames(cwf.out)[i]=="type I") {
    for (j in 1:nrow(cwf.out)) {
      if (cwf.out[j,i]!="NA" && as.numeric(cwf.out[j,i]) > 0.1) {
        cwf.out[j,i] <- paste0("\\color{lightgray} ",cwf.out[j,i])
        if (colnames(cwf.out)[i+1]=="power") {
          cwf.out[j,i+1] <- paste0("\\color{lightgray} ",cwf.out[j,i+1])
        }
      }
    }
  }
}
cwf.str.out <- print(xtable(cwf.out), include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
cwf.str.out <- gsub(
  "\\\\begin\\{tabular\\}\\{l*\\}", 
  "\\\\begin{tabular}{lrrrr@{\\\\hspace{1.8em}}rr}", cwf.str.out)
cwf.str.out <- gsub("toprule",
"toprule 
  & \\\\multicolumn{2}{c}{\\\\textsc{param.}} & \\\\multicolumn{2}{c@{\\\\hspace{1.8em}}}{\\\\textsc{smooth}} && \\\\\\\\
\\\\cmidrule(lr{.5em}){2-3}\\\\cmidrule(l{.5em}r{1.7em}){4-5}", cwf.str.out)
cwf.str.out <- gsub("table","table*",cwf.str.out)
clip <- pipe("pbcopy", "w")
write(cwf.str.out, file=clip)                               
close(clip)
```

## Different methods of significance testing

`sim.type == "complex_sigtest"`.

```{r}
make.table <- function (tbl, selectors) {
  tbl <- as.data.frame(tbl, stringsAsFactors=F)
  out <- data.frame(type.of.test="", 
                    type.1=0,
                    type.1=0,
                    type.1=0, power=0, memory=0, time=0,
                    stringsAsFactors=F)[0,]
  summarycol <- paste(gsub("(.*?)_.*", "\\1", tbl$diff.smooth.type),tbl$method,sep=",")
  for (s in 1:length(selectors)) {
    sel <- selectors[[s]]
    out[s,] <- list(sel$name,
                    tbl[tbl$data.type=="formants" & summarycol==sel$which.row,sel$columns[1]],
                    tbl[tbl$data.type=="pitch" & summarycol==sel$which.row,sel$columns[1]],
                    tbl[tbl$data.type=="simulated" & summarycol==sel$which.row & tbl$mod.comp==sel$mod.comp.sim,sel$columns[1]],
                    tbl[tbl$data.type=="simulated" & summarycol==sel$which.row & tbl$mod.comp==sel$mod.comp.sim,sel$columns[2]],
                    tbl[tbl$data.type=="simulated" & summarycol==sel$which.row & tbl$mod.comp==sel$mod.comp.sim,sel$columns[3]],
                    tbl[tbl$data.type=="simulated" & summarycol==sel$which.row & tbl$mod.comp==sel$mod.comp.sim,sel$columns[4]])
    print(out[s,])
  }
  return(out)
}

make.vis.table <- function (tbl, selectors) {
  tbl <- as.data.frame(tbl, stringsAsFactors=F)
  out <- data.frame(type.of.test="", type.1=0, power=0, type.1=0, power=0, stringsAsFactors=F)[0,]
  summarycol <- paste(tbl$diff.smooth.type,tbl$method,sep=",")
  for (s in 1:length(selectors)) {
    sel <- selectors[[s]]
    out[s,] <- list(sel$name,
                    tbl[summarycol==sel$which.row,paste(sel$which.p,"type.1",sep=".")],
                    tbl[summarycol==sel$which.row,paste(sel$which.p,"power",sep=".")],
                    tbl[summarycol==sel$which.row,"memory.med"],
                    tbl[summarycol==sel$which.row,"time.med"])
    print(out[s,])
  }
  return(out)
}


dat.cs.all <- dat[dat$sim.type=="complex_sigtest",]

# aggregating data
dat.cs.aggr <- 
  dat.cs.all[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                both.type.1=mean(par.p[error.type=="type_1"] < 0.05 | smooth.p[error.type=="type_1"] < 0.05),
                bonferroni.type.1=mean(par.p[error.type=="type_1"] < 0.025 | smooth.p[error.type=="type_1"] < 0.025),
                mod.comp.type.1=mean(mod.comp.p[error.type=="type_1"] < 0.05),
                AIC.median.type.1=median(mod.comp.AIC[error.type=="type_1"], na.rm=T),
                AIC.lower.type.1=quantile(mod.comp.AIC[error.type=="type_1"], 0.025, na.rm=T),
                AIC.upper.type.1=quantile(mod.comp.AIC[error.type=="type_1"], 0.975, na.rm=T),
                AIC.cutoff.type.1=mean(mod.comp.AIC[error.type=="type_1"] < 0, na.rm=T),
                smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                both.power=mean(par.p[error.type=="type_2"] < 0.05 | smooth.p[error.type=="type_2"] < 0.05),
                bonferroni.power=mean(par.p[error.type=="type_2"] < 0.025 | smooth.p[error.type=="type_2"] < 0.025),
                mod.comp.power=mean(mod.comp.p[error.type=="type_2"] < 0.05),
                AIC.median.power=median(mod.comp.AIC[error.type=="type_2"], na.rm=T),
                AIC.lower.power=quantile(mod.comp.AIC[error.type=="type_2"], 0.025, na.rm=T),
                AIC.upper.power=quantile(mod.comp.AIC[error.type=="type_2"], 0.975, na.rm=T),
                AIC.cutoff.power=mean(mod.comp.AIC[error.type=="type_2"] < 0, na.rm=T),
                time.med=median(time),
                memory.med=median(memory)),
                by=.(data.type,diff.smooth.type,method,mod.comp)]
kable(dat.cs.aggr)

csf.out <- make.table(dat.cs.aggr,
                     list(
                       list(name="par/smooth",
                            which.row="diff,discrete",
                            mod.comp.sim="nomodcomp",
                            columns=c("both.type.1","both.power","memory.med","time.med")),
                       list(name="par/smooth bonf.",
                            which.row="diff,discrete",
                            mod.comp.sim="nomodcomp",
                            columns=c("bonferroni.type.1","bonferroni.power","memory.med","time.med")),
                       list(name="binary smooth",
                            which.row="bin,discrete",
                            mod.comp.sim="nomodcomp",
                            columns=c("smooth.type.1","smooth.power","memory.med","time.med")),
                       list(name="LRT, ML",
                            which.row="diff,ML",
                            mod.comp.sim="modcomp",
                            columns=c("mod.comp.type.1","mod.comp.power","memory.med","time.med")),
                       list(name="LRT, fREML",
                            which.row="diff,discrete",
                            mod.comp.sim="modcomp",
                            columns=c("mod.comp.type.1","mod.comp.power","memory.med","time.med")),
                       list(name="LRT, fREML trick",
                            which.row="diff,discrete+select",
                            mod.comp.sim="modcomp",
                            columns=c("mod.comp.type.1","mod.comp.power","memory.med","time.med")),
                       list(name="AIC, fREML",
                            which.row="diff,discrete",
                            mod.comp.sim="modcomp",
                            columns=c("AIC.cutoff.type.1","AIC.cutoff.power","memory.med","time.med"))
                     ))

colnames(csf.out) <- c("significance test",
                      "type I","type I","type I","power",
                      "memory","time")
#csf.out <- as.data.frame(csf.out)
for (i in 2:5) {
  csf.out[,i] <- sprintf("%.3f", csf.out[,i])
}
for (i in 6:7) {
  csf.out[,i] <- sprintf("%.1f", csf.out[,i])
}

for (i in 2:5) {
  if (colnames(csf.out)[i]=="type I") {
    for (j in 1:nrow(csf.out)) {
      if (csf.out[j,i]!="NA" && as.numeric(csf.out[j,i]) > 0.1) {
        csf.out[j,i] <- paste0("\\color{lightgray} ",csf.out[j,i])
        if (colnames(csf.out)[i+1]=="power") {
          csf.out[j,i+1] <- paste0("\\color{lightgray} ",csf.out[j,i+1])
        }
      }
    }
  }
}

csf.str.out <- print(xtable(csf.out),include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
csf.str.out <- gsub(
  "\\\\begin\\{tabular\\}\\{l*\\}", 
  "\\\\begin{tabular}{lr@{\\\\hspace{1.8em}}r@{\\\\hspace{1.8em}}rr@{\\\\hspace{1.8em}}rr}", csf.str.out)
csf.str.out <- gsub("toprule",
"toprule 
  & \\\\textsc{r.\\\\ formants} & \\\\textsc{r.\\\\ pitch} & \\\\multicolumn{2}{c@{\\\\hspace{2.3em}}}{\\\\textsc{s.\\\\ formants}}\\\\\\\\ 
\\\\cmidrule(l{.5em}r{1.7em}){2-2}\\\\cmidrule(l{-.1em}r{1.7em}){3-3}\\\\cmidrule(l{-.1em}r{1.7em}){4-5}", csf.str.out)
clip <- pipe("pbcopy", "w")
write(csf.str.out, file=clip)                               
close(clip)
```

Report figures for AIC data.

```{r}
aics.form <- dat.cs.all[data.type=="formants" & startsWith(diff.smooth.type, "diff") & method=="discrete",]
cutoffs <- seq(-5,0,0.01)
type1s <- unlist(lapply(cutoffs, function (x) mean(aics.form$mod.comp.AIC < x)))
plot(cutoffs, type1s)
mean(aics.form$mod.comp.AIC < -0.2)

aics.pitch <- dat.cs.all[data.type=="pitch" & startsWith(diff.smooth.type, "diff") & method=="discrete",]
cutoffs <- seq(-5,0,0.01)
type1s <- unlist(lapply(cutoffs, function (x) mean(aics.pitch$mod.comp.AIC < x)))
plot(cutoffs, type1s)
mean(aics.pitch$mod.comp.AIC < -0.2)

aics.sim1 <- dat.cs.all[data.type=="simulated" & startsWith(diff.smooth.type, "diff") & method=="discrete" & error.type=="type_1",]
cutoffs <- seq(-5,0,0.01)
type1s <- unlist(lapply(cutoffs, function (x) mean(aics.sim1$mod.comp.AIC < x)))
plot(cutoffs, type1s)
mean(aics.sim1$mod.comp.AIC < -0.2)

aics.sim2 <- dat.cs.all[data.type=="simulated" & startsWith(diff.smooth.type, "diff") & method=="discrete" & error.type=="type_2",]
cutoffs <- seq(-5,0,0.01)
power <- unlist(lapply(cutoffs, function (x) mean(aics.sim2$mod.comp.AIC < x)))
plot(cutoffs, power)
mean(aics.sim2$mod.comp.AIC < -0.2)
```

```{r}
# no table for visual data; instead, ecdf plots are used

dat.cs.simulated.vis.type.1 <- dat.cs.all[method=="discrete" & error.type=="type_1" & data.type=="simulated" & mod.comp=="modcomp",]
dat.cs.simulated.vis.power <- dat.cs.all[method=="discrete" & error.type=="type_2" & data.type=="simulated" & mod.comp=="modcomp",]
reverse.ecdf <- function (x) {
  Vectorize(function (y) {mean(x >= y)}, "y")
}
csf.overlap.type.1 <- reverse.ecdf(dat.cs.simulated.vis.type.1$visual.no.overlap)(seq(0.01,1,0.01))
csf.overlap.power <- reverse.ecdf(dat.cs.simulated.vis.power$visual.no.overlap)(seq(0.01,1,0.01))
csf.diff.type.1 <- reverse.ecdf(dat.cs.simulated.vis.type.1$visual.excludes.0)(seq(0.01,1,0.01))
csf.diff.power <- reverse.ecdf(dat.cs.simulated.vis.power$visual.excludes.0)(seq(0.01,1,0.01))

pdf("../../jphon/revision_2/images/sim_visual.pdf",width=6, height=4.5)
plot(c(), c(), xlim=c(0,100), ylim=c(0,1), xlab="", ylab="", cex.axis=1.25,
     xaxt="n")
axis(1, at=c(1,20,40,60,80,100), cex.axis=1.25)
title(xlab="Cut-off %", cex.lab=1.25, font.lab=2)
title(ylab="Type I error rate / Power", cex.lab=1.25, font.lab=2)
title(main="Visual methods: Simulated F2", cex.lab=1.25, font.lab=2, cex.main=1.75)
lines(seq(1,100,1), csf.overlap.type.1, lwd=3, col="firebrick3", lty=2)
lines(seq(1,100,1), csf.overlap.power, lwd=3, col="deepskyblue4", lty=2)
lines(seq(1,100,1), csf.diff.type.1, lwd=3, col="firebrick3", lty=1)
lines(seq(1,100,1), csf.diff.power, lwd=3, col="deepskyblue4", lty=1)
text(5, 0.92, "difference, power", col="deepskyblue4", adj=0, cex=1.25)
text(2, 0.675, "overlap, power", col="deepskyblue4", adj=0, cex=1.25)
text(1, 0.2, "difference, type I", col="firebrick3", adj=0, cex=1.25)
text(3, 0.08, "overlap, type I", col="firebrick3", adj=0, cex=1.25)
abline(h=0.05, col="darkgray", lty=3, lw=2)
#legend("topright", c("overlap, type I","overlap, power", 
#                     "difference, type I", "difference, power"),
#       col=c("firebrick3","deepskyblue4","firebrick3","deepskyblue4"),
#       lty=c(2,2,1,1), lwd=rep(2,4))
dev.off()
```

```{r}
# no table for visual data; instead, ecdf plots are used

dat.cs.formants.vis.type.1 <- dat.cs.all[method=="discrete" & error.type=="type_1" & data.type=="formants" & mod.comp=="modcomp",]
dat.cs.pitch.vis.type.1 <- dat.cs.all[method=="discrete" & error.type=="type_1" & data.type=="pitch" & mod.comp=="modcomp",]

reverse.ecdf <- function (x) {
  Vectorize(function (y) {mean(x >= y)}, "y")
}
csf.overlap.type.1 <- reverse.ecdf(dat.cs.formants.vis.type.1$visual.no.overlap)(seq(0.01,1,0.01))
csp.overlap.type.1 <- reverse.ecdf(dat.cs.pitch.vis.type.1$visual.no.overlap)(seq(0.01,1,0.01))
csf.diff.type.1 <- reverse.ecdf(dat.cs.formants.vis.type.1$visual.excludes.0)(seq(0.01,1,0.01))
csp.diff.type.1 <- reverse.ecdf(dat.cs.pitch.vis.type.1$visual.excludes.0)(seq(0.01,1,0.01))

pdf("../../jphon/revision_2/images/formant_pitch_visual.pdf",width=6, height=4.5)
par(mfrow=c(2,1))
par(oma = c(0, 2, 0, 0))
par(mar=c(1.1,2.1,4.1,2.1))
plot(c(), c(), xlim=c(0,100), ylim=c(0,0.25), xlab="", ylab="", cex.axis=1.25,
     xaxt="n", yaxt="n")
axis(2, at=c(0,0.1,0.2), labels=c("0.0", "0.1", "0.2"), cex.axis=1.25)
#axis(1, at=c(1,20,40,60,80,100), cex.axis=1.25)
#title(ylab="Type I\n error rate", cex.lab=1.25, font.lab=2)
title(main="Visual methods: Real F2 / Pitch", cex.lab=1.25, font.lab=2, cex.main=1.75)
lines(seq(1,100,1), csf.overlap.type.1, lwd=3, col="firebrick3", lty=2)
lines(seq(1,100,1), csf.diff.type.1, lwd=3, col="firebrick3", lty=1)

lines(c(-13,-1), c(0.05, 0.05), col="darkgray", lty=3, lw=2)
lines(c(32,200), c(0.05, 0.05), col="darkgray", lty=3, lw=2)


text(0, 0.19, "difference, type I", col="firebrick3", adj=0, cex=1.25)
text(0, 0.06, "overlap, type I", col="firebrick3", adj=0, cex=1.25)

mtext("Real F2", side=3, line = -1.35, adj=0.98, cex=1.25, font=2)

#legend("topright", c("overlap, type I","overlap, power", 
#                     "difference, type I", "difference, power"),
#       col=c("firebrick3","deepskyblue4","firebrick3","deepskyblue4"),
#       lty=c(2,2,1,1), lwd=rep(2,4))

par(mar=c(5.1,2.1,0.1,2.1))
plot(c(), c(), xlim=c(0,100), ylim=c(0,0.25), xlab="", ylab="", cex.axis=1.25,
     xaxt="n", yaxt="n")
axis(1, at=c(1,20,40,60,80,100), cex.axis=1.25)
axis(2, at=c(0,0.1,0.2), cex.axis=1.25)
title(xlab="Cut-off %", cex.lab=1.25, font.lab=2)
#title(ylab="Type I\n error rate", cex.lab=1.25, font.lab=2)
lines(seq(1,100,1), csp.overlap.type.1, lwd=3, col="firebrick3", lty=2)
lines(seq(1,100,1), csp.diff.type.1, lwd=3, col="firebrick3", lty=1)
#text(1, 0.2, "difference, type I", col="firebrick3", adj=0, cex=1.25)
#text(3, 0.08, "overlap, type I", col="firebrick3", adj=0, cex=1.25)
#abline(h=0.05, col="darkgray", lty=3, lw=2)
lines(c(-13,-1), c(0.05, 0.05), col="darkgray", lty=3, lw=2)
lines(c(32,200), c(0.05, 0.05), col="darkgray", lty=3, lw=2)

text(0, 0.15, "difference, type I", col="firebrick3", adj=0, cex=1.25)
text(0, 0.05, "overlap, type I", col="firebrick3", adj=0, cex=1.25)

mtext('Type I error rate', side = 2, outer = TRUE, line = 1, cex=1.25, font=2)
mtext("Real pitch", side=3, line = -1.35, adj=0.98, cex=1.25, font=2)

dev.off()
```

```{r}
# no table for visual data; instead, ecdf plots are used

dat.cs.formants.vis.type.1 <- dat.cs.formants[dat.cs.formants$method=="discrete" & dat.cs.formants$error.type=="type_1",]
dat.cs.formants.vis.power <- dat.cs.formants[dat.cs.formants$method=="discrete" & dat.cs.formants$error.type=="type_2",]
reverse.ecdf <- function (x) {
  Vectorize(function (y) {mean(x >= y)}, "y")
}
csf.overlap.type.1 <- reverse.ecdf(dat.cs.formants.vis.type.1$visual.no.overlap)(seq(0.01,1,0.01))
csf.overlap.power <- reverse.ecdf(dat.cs.formants.vis.power$visual.no.overlap)(seq(0.01,1,0.01))
csf.diff.type.1 <- reverse.ecdf(dat.cs.formants.vis.type.1$visual.excludes.0)(seq(0.01,1,0.01))
csf.diff.power <- reverse.ecdf(dat.cs.formants.vis.power$visual.excludes.0)(seq(0.01,1,0.01))

pdf("../jphon/images/f2_visual.pdf",width=6, height=4.5)
plot(c(), c(), xlim=c(0,100), ylim=c(0,1), xlab="", ylab="", cex.axis=1.25,
     xaxt="n")
axis(1, at=c(1,20,40,60,80,100), cex.axis=1.25)
title(xlab="Cut-off %", cex.lab=1.25, font.lab=2)
title(ylab="Type I error rate / Power", cex.lab=1.25, font.lab=2)
title(main="Visual methods: F2 contours", cex.lab=1.25, font.lab=2, cex.main=1.75)
lines(seq(1,100,1), csf.overlap.type.1, lwd=3, col="firebrick3", lty=2)
lines(seq(1,100,1), csf.overlap.power, lwd=3, col="deepskyblue4", lty=2)
lines(seq(1,100,1), csf.diff.type.1, lwd=3, col="firebrick3", lty=1)
lines(seq(1,100,1), csf.diff.power, lwd=3, col="deepskyblue4", lty=1)
abline(h=0.05, col="gray", lty=3, lw=2)
legend("topright", c("overlap, type I","overlap, power", 
                     "difference, type I", "difference, power"),
       col=c("firebrick3","deepskyblue4","firebrick3","deepskyblue4"),
       lty=c(2,2,1,1), lwd=rep(2,4))
dev.off()


```

And now the pitch contours.

```{r}
dat.cs.pitch <- dat[dat$sim.type=="complex_sigtest" & dat$data.type=="pitch",]
# aggregating data
dat.cs.pitch.aggr <- 
  dat.cs.pitch[,.(smooth.type.1=mean(smooth.p[error.type=="type_1"] < 0.05),
                     both.type.1=mean(par.p[error.type=="type_1"] < 0.05 | smooth.p[error.type=="type_1"] < 0.05),
                     bonferroni.type.1=mean(par.p[error.type=="type_1"] < 0.025 | smooth.p[error.type=="type_1"] < 0.025),
                     mod.comp.type.1=mean(mod.comp.p[error.type=="type_1"] < 0.05),
                     smooth.power=mean(smooth.p[error.type=="type_2"] < 0.05),
                     both.power=mean(par.p[error.type=="type_2"] < 0.05 | smooth.p[error.type=="type_2"] < 0.05),
                     bonferroni.power=mean(par.p[error.type=="type_2"] < 0.025 | smooth.p[error.type=="type_2"] < 0.025),
                     mod.comp.power=mean(mod.comp.p[error.type=="type_2"] < 0.05),
                     time.med=median(time),
                     memory.med=median(memory)),
                  by=.(diff.smooth.type,method)]
kable(dat.cs.formants.aggr)

csp.out <- make.table(dat.cs.pitch.aggr,
                     list(
                       list(name="par/smooth",
                            which.row="diff_tp_15,discrete",
                            which.p="both"),
                       list(name="par/smooth bonf.",
                            which.row="diff_tp_15,discrete",
                            which.p="bonferroni"),
                       list(name="LRT, ML",
                            which.row="diff_tp_15,ML",
                            which.p="mod.comp"),
                       list(name="LRT, fREML",
                            which.row="diff_tp_15,discrete",
                            which.p="mod.comp"),
                       list(name="binary smooth",
                            which.row="bin_tp_15,discrete",
                            which.p="smooth"),
                       list(name="LRT, fREML trick",
                            which.row="diff_tp_15,discrete+select",
                            which.p="mod.comp")
                     ))

colnames(csp.out) <- c("significance test",
                      "type I","power",
                      "memory","time")
csp.out <- as.data.frame(csp.out)
csp.out[,2] <- as.character(round(csp.out[,2],3))
csp.out[,3] <- as.character(round(csp.out[,3],3))
for (j in 1:nrow(csp.out)) {
  if (as.numeric(csp.out[j,2]) > 0.15) {
    csp.out[j,2] <- paste0("\\color{lightgray} ",csp.out[j,2])
    csp.out[j,3] <- paste0("\\color{lightgray} ",csp.out[j,3])
  }
}

csp.str.out <- print(xtable(csp.out, digits=c(0,0,0,0,1,1)),include.rownames=FALSE,booktabs = TRUE,
      sanitize.colnames.function=function (x) {paste0("\\textsc{",x,"}")},
      sanitize.text.function=function (x) {x})
csp.str.out <- gsub("lllrr", "ll@{\\\\hspace{0.7em}}l@{\\\\hspace{0.7em}}r@{\\\\hspace{0.7em}}r", csp.str.out)
clip <- pipe("pbcopy", "w")
write(csp.str.out, file=clip)                               
close(clip)

dat.cs.pitch.vis.type.1 <- dat.cs.pitch[dat.cs.pitch$method=="ML" & dat.cs.pitch$error.type=="type_1",]
dat.cs.pitch.vis.power <- dat.cs.pitch[dat.cs.pitch$method=="ML" & dat.cs.pitch$error.type=="type_2",]
reverse.ecdf <- function (x) {
  Vectorize(function (y) {mean(x >= y)}, "y")
}
csp.overlap.type.1 <- reverse.ecdf(dat.cs.pitch.vis.type.1$visual.no.overlap)(seq(0.01,1,0.01))
csp.overlap.power <- reverse.ecdf(dat.cs.pitch.vis.power$visual.no.overlap)(seq(0.01,1,0.01))
csp.diff.type.1 <- reverse.ecdf(dat.cs.pitch.vis.type.1$visual.excludes.0)(seq(0.01,1,0.01))
csp.diff.power <- reverse.ecdf(dat.cs.pitch.vis.power$visual.excludes.0)(seq(0.01,1,0.01))

pdf("../jphon/images/pitch_visual.pdf",width=6, height=4.5)
plot(c(), c(), xlim=c(0,100), ylim=c(0,1), xlab="", ylab="", cex.axis=1.25,
     xaxt="n")
axis(1, at=c(1,20,40,60,80,100), cex.axis=1.25)
title(xlab="Cut-off %", cex.lab=1.25, font.lab=2)
title(ylab="Type I error rate / Power", cex.lab=1.25, font.lab=2)
title(main="Visual methods: Pitch contours", cex.lab=1.25, font.lab=2, cex.main=1.75)
lines(seq(1,100,1), csp.overlap.type.1, lwd=3, col="firebrick3", lty=2)
lines(seq(1,100,1), csp.overlap.power, lwd=3, col="deepskyblue4", lty=2)
lines(seq(1,100,1), csp.diff.type.1, lwd=3, col="firebrick3", lty=1)
lines(seq(1,100,1), csp.diff.power, lwd=3, col="deepskyblue4", lty=1)
abline(h=0.05, col="gray", lty=3, lw=2)
legend("topright", c("overlap, type I","overlap, power", 
                     "difference, type I", "difference, power"),
       col=c("firebrick3","deepskyblue4","firebrick3","deepskyblue4"),
       lty=c(2,2,1,1), lwd=rep(2,4))
dev.off()
```
