---
title: "Assembling master data sheet for GAMM Modelling Strategies paper (JPhon)"
output: html_notebook
---

```{r "setup", include=FALSE}
require("knitr")
# setting workind dir
opts_knit$set(root.dir = "~/documents/research/projects/dynamic-gam/gamm_strategies")
```

Importing relevant libraries.

```{r}
library(mgcv)
library(data.table)
library(stringr)
```

Here are three helper functions that allow us to read in data from the simulation output files and store it in a data frame. These functions need to be somewhat flexible as not all simulations return exactly the same type of data. Differences include:

- binary vs. difference smooths, which mean that some models lack a parametric difference term
- in simulations where visualisation data was collected, the name of the slots with the visualisation info are visual_no_overlap and visual_excludes_0; in other simulations these slots are typically called visual_diff and visual_both (but these are NULL valued)
- some simulations include model comparisons, while others don't

```{r}
# helper function for extracting p value from compareML output

ch.p.value.to.numeric <- function (x) {
  if (!is.na(x)) {
    if (as.character(x) == " < 2e-16") {
      return (0)
    } else {
      return(as.numeric(as.character(x)))
    }
  } else { # value of NA only when ML of nested model is lower than that of full model
    return (1)
  }
}

# helper function for extracting AIC value from compareML output

AIC.to.numeric <- function (x) {
  return(as.numeric(gsub("AIC difference: (-*[0-9]*[.][0-9]*).*", "\\1", x)))
}

# helper function for getting info from a single simulation run

extract.single.line <- function (x) { # where x is a single simulation run
  list(diff.smooth.type=x$parameters$fixed_effects,
       ranef=x$parameters$random_effects,
       AR=x$parameters$AR,
       method=x$parameters$method,
       mod.comp=x$parameters$mod_comp,
       data.density=x$parameters$dataset,
       random.smooth.k=ifelse(grepl("gamcheck", x$parameters$random_effects), extract.random.smooth.k(x$full_code), NA),
       smooth.p=x$summary$smooth[2,4],
       par.p=ifelse(nrow(x$summary$parametric) > 1, x$summary$parametric[2,4], NA),
       mod.comp.p=ifelse(!is.null(x$model_comparison), ch.p.value.to.numeric(x$model_comparison$table[[2,6]]), NA),
       mod.comp.AIC=ifelse(!is.null(x$model_comparison), AIC.to.numeric(x$model_comparison$AIC), NA),
       visual.no.overlap=ifelse(!is.null(x$visual_no_overlap),x$visual_no_overlap,NA),
       visual.excludes.0=ifelse(!is.null(x$visual_excludes_0),x$visual_excludes_0,NA),
       time=x$time[[3]],
       memory=x$memory)
}

# function for assembling data frame with info about each batch of simulations within a single iteration

analyse.batch <- function (output.item) {
  return( rbindlist(lapply(output.item, extract.single.line)) )
}

extract.random.smooth.k <- function (model_code) {
  return(as.numeric(str_match(model_code, 'bs=["]fs["][^)]*k=([0-9]*)')[2]))
}
```

Now we load the relevant data sets, process them and write the master data sheet to an rds file.

```{r}
# these are the subfolders in the gamm_modelling_strategies github repo
# dirs.to.read <- c("pitch/complex_sigtest")
dirs.to.read <- c("formants/simple/output_type_1",
                  "formants/complex_across/output_type_1",
                  "formants/complex_sigtest/output_type_1",
                  "pitch/simple/output_type_1",
                  "pitch/complex_across/output_type_1",
                  "pitch/complex_sigtest/output_type_1",
                  "simulated/simple/output_type_1",
                  "simulated/simple/output_type_2",
                  "simulated/complex_across/output_type_1",
                  "simulated/complex_across/output_type_2",
                  "simulated/complex_within/output_type_1",
                  "simulated/complex_within/output_type_2",
                  "simulated/complex_sigtest/output_type_1",
                  "simulated/complex_sigtest/output_type_2")
# adding location of type 1 / type 2 sims for each subfolder
# dirs.to.read <- unlist(lapply(c("/output_type_1","/output_type_2"), function (x) paste(dirs.to.read,x,sep="")))
# metadata for merging with loaded data sets (generated from folder names)
dirs.to.read.df <- data.table(sim.summary=dirs.to.read,
                              data.type=unlist(lapply(str_split(dirs.to.read, "/"), function (x) x[1])),
                              sim.type=unlist(lapply(str_split(dirs.to.read, "/"), function (x) x[2])),
                              error.type=unlist(lapply(str_split(dirs.to.read, "/"), function (x) substr(x[3], 8,13)))
                              )

# extremely complex code for very simple task of creating
# a shortened id for each simulation folder...
dirs.to.read.df$sim.sum.short <- with(dirs.to.read.df,
                                      paste(substr(data.type,1,1),
                                            ifelse(unlist(lapply(str_split(sim.type, "_"), function (x) length(x) > 1)),
                                                   unlist(lapply(str_split(sim.type, "_"), function (x) paste(substr(x[1],1,1), substr(x[2],1,1), sep=""))),
                                                   substr(sim.type, 1, 2)),
                                            substr(error.type, nchar(error.type), nchar(error.type)),
                                            sep="") 
)

# creating list of data tables
dats <- list()
for (dir in 1:length(dirs.to.read)) {
  t <- system.time({
  # list of files in relevant folder (these should all be of type rds)
  files.to.read <- list.files(dirs.to.read[dir], full.names=T)
  # this is where the real action happens: 
  # - files read in one by one
  # - each file contains a set of gamm summaries based on the same simulated data
  # - each of these summaries is converted to a list (inside analyse.batch)
  # - these are joined into a data table (inside analyse.batch)
  # - then all thes data tables are bound together (rbindlist)
  dats[[dir]] <- rbindlist(lapply(files.to.read, function (x) {analyse.batch(readRDS(x))}),idcol="sim.id")
  # adding metadata for simulations
  dats[[dir]] <- cbind(dirs.to.read.df[dir,], dats[[dir]])
  # adding ids for each simulated data set
  dats[[dir]]$sim.id <- paste(dirs.to.read.df[dir,]$sim.sum.short, dats[[dir]]$sim.id, sep="_")
  })
  cat(dirs.to.read[dir], "read in", t[[3]],"\n")
}
# creating master data set through rbinding list of data tables
dat <- rbindlist(dats)
# saving master data set
saveRDS(dat, "gamm_sim_master.rds")
```